---
title: "initial_analysis"
author: "Adam Sun"
date: "2019-10-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

### Functions
```{r}
rm(list=ls())
library(penalized)
library(optparse)
library(dplyr)
library(flashr)
library(devtools)
library(ebnm)
library(NNLM)

source("C:/Users/aclou/Grad School/2020 Autumn/Thesis/ts_eQTLs-master/ts_eQTLs-master/sn_spMF/readIn.R")
source("C:/Users/aclou/Grad School/2020 Autumn/Thesis/ts_eQTLs-master/ts_eQTLs-master/sn_spMF/Update_FL.R")


run_nnf = function(X_mtx, W_mtx, num_facs, max_iters = 100, penalty_L = 0.1, penalty_F = 0.1){
  # Write to table first in order to work with "readIn" function format
  write.table(X_mtx, 'C:/Users/aclou/Grad School/2020 Autumn/Thesis/ts_eQTLs-master/ts_eQTLs-master/X_sim.txt', row.names = FALSE)
  write.table(W_mtx, 'C:/Users/aclou/Grad School/2020 Autumn/Thesis/ts_eQTLs-master/ts_eQTLs-master/W_sim.txt', row.names = FALSE)

  # Define hyperparams for test
  K = num_facs
  alpha1 = penalty_L
  lambda1 = penalty_F
  inputdir='C:/Users/aclou/Grad School/2020 Autumn/Thesis/ts_eQTLs-master/ts_eQTLs-master/'
  outputdir = inputdir
  xfn=paste0(inputdir,'X_sim.txt')
  wfn=paste0(inputdir,'W_sim.txt')
  Data = readIn(K, alpha1, lambda1, xfn, wfn) 
  X = Data[['X']];
  W = Data[['W']];
  option = Data[['option']];
  option[['iter']]  = max_iters;
  Fn_basename = Data[['Fn_basename']];
  
  ## run MF to learn the factors  
  print(paste0('K=', (K), '; alpha1=', (alpha1),'; lambda1=', (lambda1)));
  Run_iter = Update_FL(X, W, option);
  FactorM = Run_iter[[1]]
  LoadingM = Run_iter[[2]]
  factor_corr = norm(cor(FactorM), 'F')
  L_sparsity = Run_iter[[3]]
  F_sparsity = Run_iter[[4]]
  print(paste0('Sparsity in Factor matrix =', (F_sparsity),'; Sparsity in L =', (L_sparsity), '; '))
  print(paste0((Run_iter[[5]]), ' factors remain; ; correlation between factors = ', (factor_corr)));
  return(list(FactorM = FactorM, LoadingM = LoadingM))
}
```

<br>

### Simulate a simple nx1 matrix first for understanding conceptually.
```{r}
set.seed(10000)

n_ = 200
p_ = 10
noise = 0.5

sparseF_sim = rbinom(n = p_, 1, prob = 0.25) 
F_sim = ifelse(sparseF_sim == 0, 0, rexp(n = p_, rate = 1)) %>% matrix(nrow = p_) # replicate F+ (non-neg), 1 factor (px1)
# dim(F_sim)
sparseL_sim = rbinom(n = n_, 1, prob = 0.25) #(nx1)
L_sim = ifelse(sparseL_sim == 0, 0, rnorm(n = n_, 0, 1)) %>% matrix() # (nx1)
E_sim = rnorm(n = n_*p_, 0, noise) %>% matrix(nrow = n_)
X_sim = L_sim%*%t(F_sim) + E_sim

W_sim = matrix(1, nrow = n_, ncol = p_)

```

<br>

### nnf. Figure out how to fix code for 1 factor (instead of 2)
```{r}
penal_L = 0.5 # vary this too
penal_F = 0.5 # vary this too
t_start = Sys.time()
nnf_out = run_nnf(X_mtx = X_sim, W_mtx = W_sim, num_facs = 2, max_iters = 20, penalty_L = penal_L, penalty_F = penal_F)
t_nnf = as.numeric(Sys.time() - t_start)
X_hat_nnf = nnf_out$LoadingM %*% t(nnf_out$FactorM)
se_nnf = sum((X_hat_nnf - X_sim)^2)
se_nnf_real = sum((X_hat_nnf - L_sim%*%t(F_sim))^2)

```

<br>

### Test flash_add_factors_from_data. Confirm pve with EVD (what is soft-thresholded SVD?)
```{r}
X_flash = flash_set_data(X_sim)
names(X_flash)
f = flash_add_factors_from_data(X_flash, K=10, backfit=FALSE)
f$pve
evd = eigen(cov(X_sim))
evd$values[1]/sum(evd$values)

```

<br>

### Backfit and plot fitted with actual
```{r}
X_backfit = flash_backfit(X_flash, f, verbose = FALSE)
X_backfit$pve
X_backfit$objective
round(X_backfit$ldf$f,5) # are numbers equal up to a sign change?
X_backfit$ldf$l[1:10] # could we have switched sign between l and f?

plot(X_backfit$fitted_values, L_sim%*%t(F_sim), ylab = "Actual", 
     main="compare fitted values with true LF'")

# Confirm that L%*%D%*%t(F) = fitted_values
sum(abs(X_backfit$ldf$l %*% X_backfit$ldf$d %*% t(X_backfit$ldf$f) - X_backfit$fitted_values))

se_eb = sum((X_backfit$fitted_values -  L_sim%*%t(F_sim))^2)
se_eb

```


<br>

### Greedy and plot fitted with actual. "ebnm_ash" seems to have lower objective, though very marginal. Why? 
```{r}
X_greedy = flash_add_greedy(X_flash, Kmax=10, verbose=FALSE)
X_greedy$pve
X_greedy$objective
round(X_greedy$ldf$f,5) # negative numbers do exist in the factors
X_greedy$ldf$l[1:10] # note results are flipped from backfit ldf$l

X_greedy_ash = flash_add_greedy(X_flash, Kmax=10, ebnm_fn="ebnm_ash",
                                verbose=FALSE)
X_greedy_ash$pve
X_greedy_ash$objective

se_eb = sum((X_greedy_ash$fitted_values -  L_sim%*%t(F_sim))^2)
se_eb

round(X_greedy_ash$ldf$f,5) # negative numbers do exist in the factors


```

<br>

### non-negative mf proposed by JW, first using greedy <br>
Note that both F and L are both non-negative <br>
https://willwerscheid.github.io/FLASHvestigations/nonnegative.html
```{r}
udv_nn = function(Y, K = 1) {
  tmp = NNLM::nnmf(Y, K, verbose = FALSE)
  return(list(d = rep(1, K), u = tmp$W, v = t(tmp$H)))
}

# One method to do non-negative mf
ebnm_fn = "ebnm_ash"
ebnm_param = list(mixcompdist = "+uniform", warmstart = TRUE)

run_flash_once <- function(f_init) {
  flash(X_sim, f_init = f_init, 
        ebnm_fn = ebnm_fn, ebnm_param = ebnm_param,
        var_type="constant", init_fn = udv_nn, 
        backfit = TRUE, verbose = FALSE)
}

fl_g <- run_flash_once(f_init = NULL)
fl_g <- run_flash_once(f_init = fl_g)
fl_g <- run_flash_once(f_init = fl_g)
fl_g$objective
fl_g$ldf$f
which(fl_g$ldf$l <= 0)

se_eb = sum((fl_g$fitted_values -  L_sim%*%t(F_sim))^2)
se_eb


```

<br>

### non-negative mf proposed by JW, using backfit. Since true K = 1, these results do worse <br>
Note again that both F and L are both non-negative <br>
```{r}
fl_nnmf <- flash_add_factors_from_data(X_flash, K = 10,
                                       init_fn = udv_nn, 
                                       backfit = FALSE)

fl_b <- flash_backfit(X_flash, fl_nnmf,
                      ebnm_fn = ebnm_fn,
                      ebnm_param = ebnm_param,
                      var_type = "constant",
                      verbose = FALSE)

fl_b$objective
which(fl_b$ldf$l <= 0)

se_eb = sum((fl_b$fitted_values -  L_sim%*%t(F_sim))^2)
se_eb

```

<br>

### How to set functions for L and F separately? <br>
Use ebnm_param to separately set the two distributions
```{r}
ebnm_fn = "ebnm_ash"
ebnm_param2 = list(l=list(), f=list(mixcompdist = "+uniform"), warmstart = TRUE)

fl_nnmf <- flash_add_factors_from_data(X_flash, K = 1,
                                       init_fn = udv_nn, 
                                       backfit = FALSE)

fl_b2 <- flash_backfit(X_flash, fl_nnmf,
                       ebnm_fn = ebnm_fn,
                       ebnm_param = ebnm_param2,
                       var_type = "constant",
                       verbose = FALSE)

fl_b2$objective

se_eb_nnL = sum((fl_b2$fitted_values -  L_sim%*%t(F_sim))^2)
se_eb_nnL

```


